\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-05 misc/JSS article/final/article.Rnw
Minor updates.

# Loading libraries and installing if unavailable
if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
# Package versions used in the article:
# fastverse 0.3.4, collapse 2.1.2, data.table 1.16.4,
# bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0

@@

# This is the benchmark function (janitor 2.2.0 is used)
bmark <- function(...) {
  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
}

@@

Since June 2024, there is also an independent \href{https://github.com/AdrianAntico/Benchmarks?tab=readme-ov-file#background}{database-like operations benchmark by Adrian Antico} using a windows server and executing scripts inside IDEs (VScode, Rstudio), on which \pkg{collapse} achieved the overall fastest runtimes. I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages. These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on personal computers. % \footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}

@@

\section*{Computational details}
The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.16.4, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.0.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.3 \citep{rbench}. All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/. The benchmark was run on an Apple MacBook Pro (2024) with 48GB M4 Pro chip. Packages were compiled from source using Homebrew Clang version 16.0.4 with OpenMP enabled and the -O2 flag. \newline

>>>>

# Loading libraries and installing if unavailable
if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
# Package versions used in the article:
# fastverse 0.3.4, collapse 2.1.2, data.table 1.17.0,
# bench 1.1.4, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.5.0
pkg <- c("bench", "dplyr", "tidyr", "matrixStats", "janitor")
if(!all(avail <- is_installed(pkg))) install.packages(pkg[!avail])

@@

# This is the benchmark function (janitor 2.2.0 is used)
bmark <- function(...) {
  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
}

@@

Since June 2024, there is also an independent \href{https://github.com/AdrianAntico/Benchmarks?tab=readme-ov-file#background}{database-like operations benchmark by Adrian Antico} using a windows server and executing scripts inside IDEs (VScode, Rstudio), on which \pkg{collapse} achieved the overall fastest runtimes. I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages. These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and personal computers. % \footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}

@@

\section*{Computational details}
The results in this paper were obtained using \proglang{R} \citep{R} 4.4.3 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.17.0, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.5.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.4 \citep{rbench}. All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/. The benchmark was run on an Apple MacBook Pro (2024) with 48GB M4 Pro chip. Packages were compiled from source using Apple Clang version 17.0.0 with OpenMP enabled and the -O2 flag. \newline
